---
title: "R Markdown for Reproducibility"
author: "Patrick Mathias"
output: html_document
---

## Setup

The first chunk in the document is typically where you load all packages you will use throughout the document.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```

## Where am I?

Use `getwd()` to show your working directory.

```{r wd}
getwd()
```

## Load a file

```{r load_file}
menu <- read_csv("data/test_menu.csv")
```

What does RStudio tell you after you import the data?

## Exercise 3

1. Locate and load the orders data set for the course from the data folder into an object (data frame) called "orders". 
2. Review the data in the new data frame using the `glimpse()` function. 
3. Within your Environment tab (top right), click the name of the object and review the raw data.

```{r}

```

```{r}
orders <- read_excel("data/orders_data_set.xlsx")
```

## Quick summary

```{r}
summary(orders)
```

## Tabulating

`tabyl` function can create quick cross tabluations to drill down into categorical fields.

```{r}
tabyl(orders, department)
```


```{r}
tabyl(orders, department, order_status_c_descr)
```

## Exercise 4

Which 3 departments ordered the highest number of labs using “Provider Preference Lists”? (pref_list_type variable)

```{r}

```

```{r}
tabyl(orders, department, pref_list_type)
```

