---
title: "Transform Data"
---

## Your Turn 1

```{r Your Turn 1, message=FALSE}
library(tidyverse)  # Provides functions used throughout this session

covid_testing <- read_csv("data/covid_testing.csv") #Read in data
```

How can you confirm that you have successfully loaded the data file into Rstudio?

	1. The code that imported the data did not yield an error
	2. Code that references `covid_testing` object runs without errors
	3. The `covid_testing` object is present in the environment pane
	4. All of the above




## select()

```{r select}
select(covid_testing, mrn, last_name)   # select columns to keep
select(covid_testing, -mrn, -last_name) # select columns to remove 
```

## Your Turn 2

A. Alter the code to select just the `first_name` column. Assign the result to `covid_testing_2a` and show the resulting data frame by executing the following line with just the name of the object: `covid_testing_2a` 

```{r Your Turn 2}
# Remove the "_______" and replace it with your answer
covid_testing_2a <- select(covid_testing, ________)
covid_testing_2a
```

B. If you have time, alter this line of code and see if you can select all columns EXCEPT the `first_name` column.

```{r Your Turn 2b}
# Remove the "_______" and replace it with your answer
covid_testing_2b <- select(covid_testing, ________)
covid_testing_2b
```


## Select Challenge

If you have extra time try the following challenge:

In the `covid_testing` dataset the turnaround time metrics are labeled `col_rec_tat` and `rec_var_tat`. Your goal is to select just these columns. Using what we learned as well as the dplyr cheatsheet which of the following code snippets could you use to select these columns?

```{r select challenge}
select(covid_testing, contains("rec"))
select(covid_testing, col_rec_tat, rec_ver_tat)
select(covid_testing, ends_with("tat"))
#all of the above
```




# line 65


## filter()

```{r}
filter(covid_testing, mrn == 5000078)
```

```{r}
filter(covid_testing, last_name=="stark")
```

## Your Turn 3

Use the 3 code blocks below, and `filter()`, to show how you would filter the `covid_testing` dataset.


A. You work in a pediatric hospital so there shouldn't too many adults in the data set. Find every test for patients over age 80. The column in which age is found is named `age`. Assign the new data frame to the object `covid_testing_3a`.
```{r Your Turn 3a}
covid_testing_3a <- filter(covid_testing, ________) #replace the ____ with a logical statement identifying tests for patients greater than 80
covid_testing_3a
```


B. Do we have a lot of outreach work for COVID testing? Find all of the covid testing where the demographic group (`demo_group`) is equal to "client". Assign the new data frame to the object `covid_testing_3b`.
```{r Your Turn 3b}
covid_testing_3b <- filter(covid_testing, ________)
covid_testing_3b
```


C. CHALLENGE: We may have some data anomolies in the data set. Find all of the covid testing where the patient class (column `patient_class`) is not entered. In these cases the value will be NA. NA stands for "not available". It is the way that R codes missing values.  [Hint: See slide titled "Logical Tests"]
```{r Your Turn 3c}

```




# line 105



## Arrange

```{r arrange}
arrange(covid_testing, first_name) #Sort in alphabetical order
arrange(covid_testing, desc(mrn))  #Sort from largest to smallest MRN
```

## Your Turn 4
The column `ct_result` contains the cycle threshold (Ct) for the real-time PCR that generated the final result.

How might you use arrange() to determine the highest and lowest Ct result in the dataset? 

```{r Your Turn 4}
arrange(covid_testing, __________)
```


## Pop Quiz

```{r pop quiz}
toy_data <- data.frame(nums=c(20,100,3),chars=c("20","100","3"))
arrange(toy_data,nums)
arrange(toy_data,chars)
```


# line 135

## Steps and the pipe

```{r Steps and the pipe}
# These 3 lines of code are an example of the how we would accomplish a multi-step analysis based on what we've learned so far. 
day_10 <- filter(covid_testing, pan_day <= 10)
day_10 <- select(day_10, clinic_name)
day_10 <- arrange(day_10 , clinic_name)

# vs

# This code accomplishes the same task and is far more legible. 
covid_testing %>% 
  filter(pan_day <= 10) %>% 
  select(clinic_name) %>% 
  arrange(clinic_name)
```

## Your Turn 5

Use `%>%` in the following chunk to write a sequence of functions that: 

1. Filters to tests from the clinic (`clinic_name`) of "picu"

2. Selects the column with the receive to verify turnaround time (`rec_ver_tat`) as well as the day from start of the pandemic (`pan_day`)

3. Arrange the `pan_day` from highest to lowest so you can see most recent tests on top.

Using "<-", assign the result to a new variable, call it whatever you want.

CHALLENGE: Use your mouse to select the name of the new data frame from the list of objects in the Environment tab in the upper-right pane of Rstudio. Answer the following questions:

- Can you find the test that exceeded 10 hours in TAT?
- What is the typical TAT for the PICU?
- Given this data does it appear that there is a chronic problem in TAT for the PICU, or was this test an outlier?

```{r Your Turn 5}
covid_testing %>% 
  filter(________) %>% #Here filter to the clinic with the name "picu"
  select(________) %>% #Here select the columns you want to keep
  arrange(________)    #Here arrange the pan_day column from highest to lowest (check slide 42 the arrange section of this file to use the desc() function)
```



## mutate

```{r}
# The following code creates a data frame with a new column called c_r_tat_mins which contains the collect to receive TAT quantified in minutes.
covid_testing %>%
	mutate(c_r_tat_mins = col_rec_tat*60)
```

## Your Turn 6
Create a new column using the mutate() function that contains the total TAT. 
Hint: The total TAT is the sum of `col_rec_tat` and `rec_ver_tat`

```{r Your Turn 6}
covid_testing %>%
	mutate(total_tat = _________)
```


## Replace a column with `mutate()`

```{r}
covid_testing %>%
	mutate(mrn = as.character(mrn))
```


# Take aways

* Extract variables with `select()`  
* Extract cases with `filter()`  
* Arrange cases, with `arrange()`  
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  

