---
title: "Transform Data"
---

# Import Data

## Your turn 1

```{r Your Turn 1}
library(tidyverse)  # Provides functions used throughout this session

covid_testing <- read_csv("data/covid_testing.csv")
```

How can you confirm that you have successfully loaded the data file into Rstudio?

	1. The code that imported the data did not yield an error
	2. Code that references `covid_testing` object runs without errors
	3. The `covid_testing` object is present in the environment pane
	4. All of the above




## select()

```{r select}
select(covid_testing, mrn, last_name)
select(covid_testing, c(1,3))
```

## Your turn 2

1. Alter the code to select just the `first_name` column.
2. If you have time, see if you can select `first_name` using a numeric index

```{r Your Turn 2}
# Remove the "_______" and replace it with your answer
select(covid_testing, ________)
```

## Your Turn 3

The turnaround time metrics are found in columns 16 and 17 and are labeled `col_rec_tat` and `rec_var_tat`. Which of the following code snippets could you use to remove these columns?

```{r Your Turn 3}
select(covid_testing, -16,-17)
select(covid_testing, -col_rec_tat, -rec_ver_tat)
select(covid_testing, -contains("tat"))
#all of the above
```



# line 55


## filter()

```{r}
filter(covid_testing, mrn == 5000078)
```

```{r}
filter(covid_testing, last_name=="stark")
```

## Your turn 4

Use the 3 code blocks below, and `filter()`, to show how you would filter the `covid_testing` dataset.

Modify the code in this block to find:

1. Every test for patients over age 80
```{r Your Turn 4a}
filter(covid_testing, ________)
```

2. All of the covid testing where the demographic group (demo_group) is equal to "client"
```{r Your Turn 4b}

```

3. CHALLENGE: All of the covid testing where the patient class (patient_class) is NA [Hint: See slide titled "Logical Tests"]
```{r Your Turn 4c}

```


# line 90



## Arrange

```{r arrange}
arrange(covid_testing, patient_id)
```

## Your Turn 5
The column `ct_result` contains the cycle threshold (Ct) for the real-time PCR that generated the final result.

How might you use arrange() to determine the highest and lowest Ct result in the dataset? 

```{r Your Turn 5}
arrange(covid_testing, __________)
```


## Pop Quiz

```{r pop quiz}
toy_data <- data.frame(nums=c(20,100,3),chars=c("20","100","3"))
arrange(toy_data,nums)
arrange(toy_data,chars)
```



# line 120

## Steps and the pipe

```{r Steps and the pipe}

day_10 <- filter(covid_testing, pan_day <= 10)
day_10 <- select(day_10, clinic_name)
day_10 <- arrange(day_10 , clinic_name)

# vs

covid_testing %>% 
  filter(pan_day <= 10) %>% 
  select(clinic_name) %>% 
  arrange(clinic_name)
```

## Your turn 6

Use `%>%` to write a sequence of functions that: 

1. Filters to tests not from the clinic (`clinic_name`) of "picu"

2. Selects the column with the receive to verify turnaround time (`rec_ver_tat`) as well as the day from start of the pandemic (`pan_day`)

3. Arrange the `pan_day` from highest to lowest

Using <-, assign the result to a new variable, call it whatever you want.

CHALLENGE: Use your mouse to select the name of the new data frame from the list of data sets in the upper-right pane of Rstudio. What is the typical TAT for the PICU?

```{r Your Turn 6}
covid_testing %>% 
  filter() %>% 
  select() %>% 
  arrange()
```







- line 165

## mutate

```{r}
covid_testing %>%
	mutate(total_tat = col_rec_tat*60)
```

## Your turn 7

Create a new column using the mutate() function that contains the total TAT. 
Hint: The total TAT is the sum of `col_rec_tat` and `rec_ver_tat`

```{r Your Turn 7}

```


## Replace a column with `mutate()`

```{r}
covid_testing %>%
	mutate(mrn = as.character(mrn))
```


# Take aways

* Extract variables with `select()`  
* Extract cases with `filter()`  
* Arrange cases, with `arrange()`  
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  

