---
title: "Visualize Data"
output: html_notebook
---

### Import Data

Run the following code to import the practice data set. 

```{r}
library(tidyverse)
library(readxl)
orders <- read_excel("data/orders_data_set.xlsx")
```

### Your Turn 4

Let's recreate the scatterplot of *result_time* vs *order_time*. 

You will need to:

1. Pick a data set: *orders*
2. Choose a geom function: *geom_point()*
3. Map aesthetic properties of the graph to variables of the data frame (in this case, the *order_time* variable should be on the x axis, and *result_time* should be on the y axis)

Fill in the blanks and run the code to create the plot.

(Hint: if you get stuck, take a look at the slide titled "ggplot()"!)

```{r}
ggplot(data = ______) +
  ______(mapping = aes(______))
```

```{r}
# Solution
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, y = result_time))
```

Let's add some color, shall we?

Copy and paste the previous code chunk and add the following aesthetic mapping: color = order_class_c_descr

```{r, message = FALSE}
# Solution
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, 
                           y = result_time, 
                           color = order_class_c_descr))
```

What happens if you additionally map the *shape* aesthetic to order_class_c_descr? Copy and paste the previous code chunk below and additionally map shape = order_class_c_descr. 

```{r, message = FALSE}
# Solution
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, 
                           y = result_time, 
                           color = order_class_c_descr,
                           shape = order_class_c_descr))
```

Stop here.
**********

