---
title: "Transform Data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r setup}
library(tidyverse) 
library(readxl) 

orders <- read_excel("data/orders_data_set.xlsx")
```

## Summarize

```{r}
orders %>%
  select(order_id, patient_id) %>%
  head(4) %>%
	summarize(order_count = n(),
	          pt_count    = n_distinct(patient_id)) 
```

## Your Turn 1

Add onto the code in the above chunk to calculate:

1) Mean count of orders per patient
2) Mean count of orders per department

*Hints:* 

- Start by adding a new code block here (copy/paste above or CTRL-ALT-I).
- Then copy code from the previous code block and edit to determine the mean number of orders per patient. 
- Next, add code to calculate the mean count of orders per department. 
- Assign the results to a new object, orders_1.

```{r}
# solution
orders_1 <- orders %>%
	summarize(order_count = n(),
	          pt_count = n_distinct(patient_id),
	          dept_count = n_distinct(department)) %>%
	mutate(pt_order_count_mean = order_count / pt_count,
         dept_order_count_mean = order_count / dept_count)
```

## Your Turn 2

Use summarize() to calculate:

1) The date of the first (or minimum) order
2) The median time difference between order_time and result_time

*Hint: Refer to help for NA handling*

```{r}
#solution
orders %>% 
	mutate(result_interval = result_time - order_time) %>%
	summarize(order_time_min = min(order_time),
		        order_time_first = first(order_time), # assumes sorted observations
            result_interval_med = median(result_interval, na.rm = T))
```

## group_by

```{r}
orders %>% 
  group_by(patient_id, department) %>%
  head(4)
```

## group_by %>% summarize

```{r}
orders %>%
	group_by(patient_id) %>%
	summarize(order_count = n()) %>%
	summarize(pt_order_count_mean = mean(order_count))
```

## Your Turn 4

1) Calculate the median number of orders per patient
2) Calculate the maximum number of TSH orders per patient
3) The 5th and 95th percentile of the number of orders per patient

```{r}
# solution
orders %>%
	group_by(patient_id) %>%
	summarize(pt_order_count = n()) %>%
	summarize(pt_order_count_median = median(pt_order_count))
```

```{r}
#solution
orders %>%  
	filter(proc_code == "TSH") %>%	
	group_by(patient_id) %>%
	summarize(pt_order_count = n()) %>%
	summarize(pt_order_count_max = max(pt_order_count))
```

```{r}
#solution
orders %>%
	group_by(patient_id) %>%
	summarize(pt_order_count = n()) %>%
	summarize(pt_order_count_q05 = quantile(pt_order_count, probs=0.05),  
	          pt_order_count_q95 = quantile(pt_order_count, probs=0.95))
```

## Your Turn 5
Calculate the mean order count per patient for each department

*Hint: summarize() rolls up a single grouping variable at a time*

```{r}
#solution
orders %>%	    
	group_by(department, patient_id) %>%	    
	summarize(order_count = n()) %>%    # summarize over department-patient_id groups
	summarize(pt_order_count_mean = mean(order_count)) # summarize over department groups
```

## Compare two sets of count data

- Is the distribution of the count of orders per patient different in the Internal Medicine and Family Medicine clinics?

## Create sample distributions

```{r}
orders_per_pt_dept <- orders %>%	        
				group_by(department, patient_id) %>%	        
				summarize(order_count = n()) %>%     
				ungroup() %>%        
				filter(department %in% 
					c("INTERNAL MEDICINE CLINIC", 
						"FAMILY MEDICINE CLINIC"))
head(orders_per_pt_dept)
```

## Visualize the distribution of order counts

```{r}
orders_per_pt_dept %>%
	ggplot() +
	geom_histogram(aes(x = order_count))
```

## Describe the central tendency of the order count

```{r}
orders_per_pt_dept %>% 
		group_by(department) %>% 
		summarize(order_count_median = median(order_count), 
			        order_count_mean = mean(order_count))
```

## Compare the distributions of order counts

```{r}
wilcox.test(order_count ~ department, 
            data        = orders_per_pt_dept, 
            alternative = "two.sided", 
            paired      = FALSE, 
            conf.int    = TRUE)
```

## Your Turn 6

Compare order counts across the nephrology, cardiology, and gastroenterology departments. Are they different?

*Hint: Convert department to a factor using function factor()*

```{r}
#solution
orders %>%	            
	group_by(department, patient_id) %>%	            
	summarize(order_count = n()) %>%     
	ungroup() %>%    
	filter(department %in% c("NEPHROLOGY CLINIC",
	                         "CARDIOLOGY CLINIC",
	                         "GASTROENTEROLOGY CLINIC")) %>%
	mutate(department = factor(department)) %>%    
	kruskal.test(order_count ~ department, 
	             data = .)
```

