---
title: "06 - Exploratory Data Analysis"
output: html_notebook
---

## Overview of Exploratory Data Analysis

Exploratory data analysis (EDA) is a process used to gain better understanding of a data set. The goal is to gain familiarity with the shape and contents of the data. The process often generates questions that can be further addressed through the use of statisical models and graphs. It is an iterative process that generally involves creating and exploring one or two dimensional summaries of the data that lead to the creation of additional summaries and ultimately an understanding of the data set. It also plays a role in data quality control, an important process by which the integrity of the data is evaluated prior to any decision-making or conclusions are made.

In this lesson we will gain more experience with some of the tools we have discussed through an EDA exercise. For many of the questions we will ask you to answer there is no right or wrong way to answer the question. However, this is an opportunity to further commit some of the new functions you have learned to memory.

## Introduction to data set

In this exercise we will be exploring a snippet of the MIMIC data set. MIMIC is an openly available dataset developed by the MIT Lab for Computational Physiology, comprising deidentified health data associated with ~40,000 critical care patients. It includes demographics, vital signs, laboratory tests, medications, and more. See this publication for additional information:

> MIMIC-III, a freely accessible critical care database. Johnson AEW, Pollard TJ, Shen L, Lehman L, Feng M, Ghassemi M, Moody B, Szolovits P, Celi LA, and Mark RG. Scientific Data (2016). DOI: 10.1038/sdata.2016.35. Available at: http://www.nature.com/articles/sdata201635

The subset that we will be exploring is a single table which includes laboratory data and some demographic information.


## Exercise 0

Before we begin we have to load our library of functions that we'll be using. Since all of the necessary functions are included in the tidyverse family of packages, all we need to do is run the library() command on tidyverse.

- Use the code block below to run `library()` on "tidyverse"

```{r}
library(tidyverse)
```


## Exercise 1

So far so good. Now we need to get our data into the rstudio environment so that we can start poking at it. Relative to our current working directory the data is located in a subfolder called "data". The file is called "mimic.csv". 

- In the code block below use the `read_csv()` function to read this data file into a new data frame called "mimic"
- Take a look at the data frame. Use `summary()`. Press on the name of the data frame in the environment tab on the upper-right.

**HINT** check out how we loaded the orders data set in the second and third lessons!

```{r}
mimic <- read_csv("data/mimic.csv")
```



