---
title: "Visualize Data - Solutions"
output: html_notebook
---

### Your Turn 4: The ggplot template and aesthetics

Run the following code chunk to set up the practice data sets.

```{r, warning = FALSE}
library(tidyverse)
library(readxl)
library(lubridate)

orders <- read_excel("data/orders_data_set.xlsx") 

a1c <- orders %>%
  mutate(order_to_result = difftime(result_time, order_time, unit = "days")) %>%
  filter(proc_code == "A1C")
```




Let's recreate the scatterplot of *result_time* vs *order_time*. 

You will need to:

1. Pick a data set: *orders*
2. Choose a geom function: *geom_point()*
3. Map aesthetic properties of the graph to variables of the data frame (in this case, the *order_time* variable should be on the x axis, and *result_time* should be on the y axis)

Fill in the blanks and run the code to create the plot.

(Hint: if you get stuck, take a look at the slide titled "ggplot()"!)

```{r}
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, y = result_time))
```




Let's add some color, shall we?

Copy and paste the previous code chunk and add the following aesthetic mapping: 

    color = order_class_c_descr

```{r, message = FALSE}
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, 
                           y = result_time, 
                           color = order_class_c_descr))
```




What happens if you additionally map the *shape* aesthetic to order_class_c_descr? Copy and paste the previous code chunk below and additionally map:

    shape = order_class_c_descr

```{r, message = FALSE}
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, 
                           y = result_time, 
                           color = order_class_c_descr,
                           shape = order_class_c_descr))
```

Stop here.

*******************************************************





### Your Turn 5: Geom functions

Pair up.

Run the following code chunk:

```{r}
ggplot(data = a1c) + 
  geom_point(mapping = aes(x = order_status_c_descr, y = order_to_result))
```




Now, with your partner, figure out how you would modify the code above so it draws boxplots instead of scatterplots. Use the cheatsheet!

```{r}
ggplot(data = a1c) + 
  geom_boxplot(mapping = aes(x = order_status_c_descr, y = order_to_result))
```



With your partner, make a histogram of the order_to_result variable. Hint: do not map the y aesthetic! In a histogram, y is always the count of values falling into a specific bin.

```{r}
ggplot(data = a1c) + 
  geom_histogram(mapping = aes(x = order_to_result))
```




When you run this code, you will most likely get what looks like an error:

  `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

This is actually just a message from R, telling you that it's a bad idea to use a default bin width. 

The order_to_result variable holds the time, in days, from when the order was placed to when the result was reported in the laboratory information system. What would be a reasonable binwidth for order_to_result? Discuss with your partner. Then fill in the code below:

```{r}
ggplot(data = a1c) + 
  geom_histogram(mapping = aes(x = order_to_result), binwidth = 1)
```




With your partner, predict what the following code will do. 

Then run it.

```{r}
ggplot(data = orders) +
  geom_point(mapping = aes(x = order_time, y = result_time)) +
  geom_smooth(mapping = aes(x = order_time, y = result_time))
```

